TBD

### 2020-04-16 Articles/2019-Entropy/Source/entropy-article.tex
Another pass at the process domain structure

The general idea is that we need to describe a system as it evolves in time. To that end, we need to define what we mean by states, processes and evolutions in the most general setting such that it can be applied to thermodyamic, dynamical, classical, quantum systems alike.

We call \textbf{process domain} $\tdomain[P]$ the set of all possible descriptions of the system. Depending on the system, we will have statements like \statement{the average volume between 1 and 2 seconds is between 3 and 4 liters}, \statement{the half-life of the material is between 4.46 and 4.48 billion years} or \statement{at time 2.5 s the position of the ball is [3.1, 4.2, -1.2] meters}. We take these statements to form a logical system, which means $\sigma$-complete Boolean algebra, that is $\tdomain[P]$ is closed under negation (NOT), countable disjunction (OR) and countable conjunction (AND).

We call \textbf{experimental domain} for the process the subset $\edomain[P] \subseteq \tdomain[P]$ of statements that can be, at least in line of principle, experimentally verified. Note that this will not contain, in general, all statements. For example, infinite precision statements like \statement{the acceleration is exactly 9.8066543621 $m/s^2$} cannot be verified.\footnote{As discussed further in the appendix, since experimental verification is constrained by finite time and finite resources, $\edomain[P]$ will not be a $\sigma$-complete algebra but, rather, it will be a Heyting algebra: it will only be closed under finite conjunction (AND) and countable disjuction (OR). Because our process must be fully defined by physically meaningful statements, we will have that $\tdomain[P]$ will be the closure of $\edomain[P]$ over negation and countable disjunction. We also require $\edomain[P]$ to have a countable basis so that the each statement in the domain can, if given enough time, verified.}

We call \textbf{evolution} a statement in $\tdomain[P]$ that, if assumed to be true, will define the truth value for all other statements. For example, if our process describes the vertical motion of a point-particle and we posit \statement{the trajectory of the particle is $y=10 \, m - t \cdot 9.80665 \, m/s^2 $} to be true, we would know the truth value of all possible finite and infinite precision measurements at all times. A process will therefore define a set of \textbf{possible evolutions}, noted $E \subset \tdomain$, which characterizes what evolutions are physically allowed. For example, in a relativistic setting evolutions with velocity greater than the speed of light would not be allowed. The possible evolutions, then, will put constraints on the system in terms of logical relationship between statements at different times: if something is true at one time then something else will be true at another time.

It is useful, at this point, to at least mention two types of relationships we can have. Given two statements $\stmt_1, \stmt_2 \in \tdomain[P]$ we say one is \textbf{narrower} than the other, noted $\stmt_1 \narrower \stmt_2$, if the second is true whenever the first is. This can describe relationships at different scale (e.g. \statement{the horizontal position is between 2.5 and 3 meters} $\narrower$ \statement{the horizontal position is between 2 and 3.5 meters}), on different quantities (e.g. \statement{the horizontal position is between 2.5 and 3 meters and the vertical position is between 1 and 1.5 meters} $\narrower$ \statement{the horizontal position is between 2.5 and 3 meters}), constraints between variables at the same time (e.g. \statement{the temperature of the water in the glass is 3.98 C} $\narrower$ \statement{the density of the water in the glass is 1 $g/cm^3$}) or at different times (e.g. \statement{at time 0 s the position is 1 m and the velocity is 1 m/s} $\narrower$ \statement{at time 1 s the position is 2 m}). Similarly, we say that two statements are \textbf{compatible}, noted $\stmt_1 \comp \stmt_2$, if they can both be true or incompatible if the cannot. This and other relationships can be defined to describe what ultimately will be the equations of state and laws of evolution that govern the process.

For each statement $\stmt \in \tdomain[P]$ we can now define the set $E(\stmt) \subseteq E$ of all the evolutions that are compatible. For a point-particle, the evolutions compatible with \statement{at time 0 s the position is 1 m} correspond exactly to all trajectories that pass at the given position at the given time. As the process domain $\tdomain[P]$ is a $\sigma$-complete Boolean algebra, the set of all subsets given by $E(\tdomain[P])$ will be $\sigma$-algebra over $E$.\footnote{The experimental domain will instead induce a topology over $E$. This can be used as a foundation to extend the framework to include geometrical notions that are outside the scope of this article.} If we are studying a random process, we can think of $E$ as the sample space and $E(\tdomain[P])$ as the event space.

As one can see from the examples, the properties and quantities for the system typically become ways to label the statements. Yet, they do not form the basis of the basic structure and, in fact, we will make no use of them. All our conclusions, then, will only depend on the idea that we have a process that we study that is describable in terms of experimentally verifiable assertions. It does not matter what specific quantities are used, or whether the process is from physics, chemistry, control systems, ecology, economics, sociology, or some other realm we have yet to discover.

>>>>

The general idea is that we need to describe a system as it evolves in time. To that end, we need to define what we mean by states, processes and evolutions in the most general setting such that it can be applied to thermodyamic, dynamical, classical, quantum systems alike. A formal and more rigorous introduction of the framework, both mathematically and conceptually, is given in the appendix. Here we limit ourselves to a quick overview of the main point. We will we first attempt to use measure theory, which hopefully the reader is already familiar with, show its shortcomings and then present our framework as an amendment.

To define a measure space we start with a set of objects that fully define all variables.\footnote{In probability theory this is called the sample space.} In our case, it will be the set $E$ of \textbf{possible evolutions} which represent full descriptions of the system at all times. For example, if our process describes the vertical motion of a point-particle, an evolution would be something like \statement{the trajectory of the particle is $y=10 \, m - t \cdot 9.80665 \, m/s^2 $}. Through this set, the process implicitly characterizes what is physically allowed. For example, in a relativistic setting evolutions with velocity greater than the speed of light would not be allowed and therefore no be part of $E$.

Next we need a $\sigma$-algebra on $E$ which can be thought as a set of incomplete descriptions about the system.\footnote{In probability theory this is called the set of events.} For a point-particle, the description \statement{at time 0 s the position is 1 m} would correspond to all evolutions such that the particle is at the given position at the given time. We call \textbf{process domain} $\tdomain[P]$ the set of all these descriptions as it captures all those descriptions pertinent to the process at hand.

Finally we a measure $\mu : \tdomain[P] \to \mathbb{R}$ that allows us to give a size to each description. This allows us to characterize which description is more or less precise: \statement{at time 0 s the particle is between 0 m and 1 m} is doubly more specific than \statement{at time 0 s the particle is between 0 m and 2 m} precisely because it correspond to half the evolutions of the second.

This measure theoretic view gives a sense of what we would like to do, except it doesn't work and needs to be amended. Its biggest strength is its generality. Even though we typically use properties and quantities for the system to label the statements, they do not appear in the basic structure. Therefore, depending on the system, we will have statements like \statement{the average volume between 1 and 2 seconds is between 3 and 4 liters}, \statement{the half-life of the material is between 4.46 and 4.48 billion years} or \statement{at time 2.5 s the position of the ball is [3.1, 4.2, -1.2] meters} which will correspond to the appropriate set of evolutions. The problem is that it fails to handle different level of descriptions in two ways.

First of all, the description given by the evolutions are mathematically different objects from all others: the first are points and the other are sets. This is awkward because evolutions of a coarse grain process will be simple descriptions on a fine grain one. So we can't properly see one as a mathematical structure contained in the other.

Second, the measure does not have enough resolution: it can only compare objects of the same dimensionality. If we imagine all evolutions to be points on an $n$ dimensional manifold, if the measure gives finite values for $k$-dimensional regions it will necessarily give measure zero for all regions with lesser dimensionality and infinite measure for those with greater dimensionality. Therefore we will not be able to relate the measure at different level of scale. This, again, is a problem since states will have different dimensionality when studying coarse and fine grained descriptions.

To solve the first problem, we note both the evolutions in $E$ and the elements of $\tdomain[P]$ are, in the end, descriptions of the system. Therefore we treat as what they are: statements in a well-posed logical system. Our process domain $\tdomain[P]$, then, is simply a set of statements that is closed under negation (NOT), countable disjunction (OR) and countable conjunction (AND), a $\sigma$-complete Boolean algebra. The evolutions $E$ are also descriptions about the system, so we simply have $E \subset \tdomain[P]$. All descriptions of subsystems or of the system at lower scale will simply be sub-algebras of $\tdomain[P]$.

All the set operations one would have can be translated in terms of logical operations. In the same way that on sets we have an inclusion operator $\subseteq$, given two statements $\stmt_1, \stmt_2 \in \tdomain[P]$ we say one is \textbf{narrower} than the other, noted $\stmt_1 \narrower \stmt_2$, if the second is true whenever the first is. This can describe relationships at different scale (e.g. \statement{the horizontal position is between 2.5 and 3 meters} $\narrower$ \statement{the horizontal position is between 2 and 3.5 meters}), on different quantities (e.g. \statement{the horizontal position is between 2.5 and 3 meters and the vertical position is between 1 and 1.5 meters} $\narrower$ \statement{the horizontal position is between 2.5 and 3 meters}), constraints between variables at the same time (e.g. \statement{the temperature of the water in the glass is 3.98 C} $\narrower$ \statement{the density of the water in the glass is 1 $g/cm^3$}) or at different times (e.g. \statement{at time 0 s the position is 1 m and the velocity is 1 m/s} $\narrower$ \statement{at time 1 s the position is 2 m}). Evolutions, then, can be seen as the narrowest statements in the domain that can still be true.

Similarly, we say that two statements are \textbf{compatible}, noted $\stmt_1 \comp \stmt_2$, if they can both be true or incompatible if the cannot, which in terms of set would correspond to the non-empty intersection. For each $\stmt \in \tdomain[P]$ we can define $E(\stmt)$ as the set of evolutions that are compatible with it. These are exactly the sets we had in the $\sigma$-algebra before, which now corresponds to $E(\tdomain[P])$. There are operations, all with straight-forward physical meaning, that can be introduced though we will stop here since these are the only two we strictly require.

This small change has significant advantages. The practical one is that we can now mix and match descriptions at different scale and of different parts of the system within a single framework. The conceptual advantage is that it is clear that all we are talking about is descriptions about a system within a model and their logical relationships, which makes it straight-forward to understand our conclusions. Lastly, the structure is still very similar to the standard tools of measure theory, therefore, once a scale or subsystem is chose, measure structures can be recovered.

The second problem is solved by replacing the measure with a suitable partial order $\finer$ on $\tdomain[P]$ which, given two statements, it tells us whether the description of one is \textbf{finer}, more refined, than the other. Suitable means, for example, that it needs to be compatible with narrowness: if $\stmt_1 \narrower\stmt_2$ we must also have $\stmt_1 \finer \stmt_2$. Two statements are \textbf{equigranualar}, noted $\stmt_1 \eqgran \stmt_2$, if they provide the same level of description. By analogy with geometrical objects, we can see that these relationships can now span different dimensionality: isolated points are always going to be smaller than lines which are always smaller than surfaces and so on, while we can still have lines and surfaces of the same size. For example, we can say that \statement{the trajectory of the particle is $y=t \cdot 1 \, m/s$} $\eqgran$ \statement{the trajectory of the particle is $y=t \cdot 2 \, m/s$} and that \statement{the position at time 0 sec is between 0 and 1 m} $\eqgran$ \statement{the position at time 0 sec is between 1 and 2 m}.

To recover measures, the idea is that, given a statement $\stmt[u] \in \tdomain[P]$ we can construct a measure $\mu_{\stmt[u]}$ such that $\mu_{\stmt[u]}(\stmt[u]) = 1$ and $\stmt_1 \finer \stmt_2$ will mean $\mu_{\stmt[u]}(\stmt_1) \leq \mu_{\stmt[u]}(\stmt_2)$. That is, we can quantity how more or less precise a statement is compared to a fixed statement taken as a unit.

This does not work in general, though, because statements along different dimensions cannot be compared. For example, consider the pressure/volume state space for an ideal gas. Comparing \statement{the pressure is 1 kPa and the volume is between 1 and 2 liters} and \statement{the pressure is between 1 and 2 kPa and the volume is 1 liter} would mean saying where one unit liter is bigger or smaller than 1 kPa. There are two types of statements in $\tdomain[P]$, though, that can always be compared.

For the first type, note that all evolutions in $E$ must give a complete description, so they must be all equigranular. Therefore any statement that corresponds to finitely many evolutions (e.g. \statement{the full trajectories is either $y=t \cdot 1 m/s$ or $y=-t \cdot 1 m/s$}) will be comparable to all other statements.

For the second group, note that in our ideal gas example one of the quantities was determined with infinite precision. If we gave a range for all quantities, we would compare areas in that space, which we can always do. Note that statements with finite precision measurements are the only ones we can experimentally verify. So, we will require that all experimentally verifiable statements are comparable. We call \textbf{experimental domain} for the process the subset $\edomain[P] \subseteq \tdomain[P]$ of statements that can be, at least in line of principle, experimentally verified.

We leave the full mathematical characterization in the mathematical section. We limit ourselves to make a few important comments. Verifiable statements will correspond, in terms of sets of $E$, to finite and infinite volumes. Requiring those to be comparable means the $E$ is a measurable space. In the familiar spaces, such as phase space or physical space, the measure we have will be the familiar one. As the process domain corresponds to a $\sigma$-algebra on $E$, the experimental domain corresponds to a topology on $E$. Therefore notions of continuity can be recovered as well. The process domain is the closure of the experimental domain under standard Boolean operations. This means that all the process domain is fully specify by, at least in principle, experimentally verifiable statements. The space does not contain any assertion that cannot ultimately be defined in terms of experiments. Lastly, requiring that, given enough time, each statement will be verified means the experimental domain must be generated by countably many statements. This means the topology is second countable and one is limited to work with ``nice'' spaces (e.g. spaces of continuous functions, separable Hilbert spaces, ...) not because they are ``nice'', but because they are the only physically meaningful ones.

To sum up, we have two foundational object. The first object is the process domain $\tdomain[P]$ which is a Boolean algebra that contain all the possible description of our process. A subset $\edomain[P]$ contains those that are experimentally verifiable, which contain all the information in the full domain. The narrowest statements in $\tdomain[P]$ correspond to the evolutions $E$, that provide the full description at all times. The second object is the operator $\finer$ which gives us a way to compare two statements and decide which one gives a more refined description. Though two statements will not in general be comparable, all verifiable statements are comparable and all evolutions provide an equally refined description.

This framework imposes very few requirements which are of very general applicability. It does not matter what specific quantities are used, or whether the process is from physics, chemistry, control systems, ecology, economics, sociology, or some other realm we have yet to discover. In the end, if we have a process we need to model, our theory will need at least to provide those objects. This means that the results we find will have broad applicability as well.
